---
title: "Movielens Capstone Project"
author: "Agnes Munee"
date: "1/20/2022"
output: pdf_document
---



## This project is based on the data Movielens, which aims to make movie recommendatuons to viewers/users based on the movies they have already watched and the rating that they gave to those movies.The features or variables under consideration are the UserID( which is the user identification), the movieId ( which is the movie Identification), the title of the movie and the genre of each movie.The "rating" is the target variable that we will be predicting.First, we analysed the data, and derived the key relationships between the variables and the target variable. This was visualized in diagramatic plots. The strength of the mentioned relationships was also measured using the correlation function. Second, we tried a couple of models, which we have descrbed in the Modelling & Results section of this report. Finally, we tested the optimal model on the validation data to determine the final RMSE.


## Method and Analysis

    # Create edx set, validation set (final hold-out test set)- (provided by the edx team)
```{r,echo=FALSE}
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)
```


```{r,echo=FALSE}
dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")


movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                            title = as.character(title),
                                            genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")


set.seed(1) 
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]


validation <- temp %>% 
      semi_join(edx, by = "movieId") %>%
      semi_join(edx, by = "userId")


removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

  

   #Check if the data has been loaded correctly
```{r,echo=FALSE}
dim(edx)
dim(validation)
```
   #Data Exploration 
   #Split the edx data into train and test data: 
```{r,echo=FALSE}
set.seed(1)
index<-createDataPartition(validation$rating,p=0.1,times = 1,list = FALSE)
test<-edx[index,]
train<-edx[-index,]

```
 #Explore the train data set
```{r,echo=FALSE}
str(train)
summary(train)
```
 #movieId
```{r,echo=FALSE}
mu<-mean(train$rating)
train%>%group_by(movieId)%>%summarize(n=n(),mean_rating=mean(rating))%>%arrange(desc(mean_rating))%>%slice(1:20)
train%>%group_by(movieId)%>%summarize(n=n(),mean_rating=mean(rating))%>%arrange(mean_rating)%>%slice(1:20)
library(ggplot2)
train%>%group_by(movieId)%>%summarize(n=n(),mean_rating=mean(rating))%>%ggplot(aes(mean_rating,n)) + geom_point()
```
The scatter plot above shows that the movies with the highest and lowest  mean rating have been rated very few times.
Thus, it would be wise to penalize the rating with the lowest number of ratings.This will be referred to as regularization.

 #userId
```{r,echo=FALSE}
train%>%group_by(userId)%>%summarize(n=n(),mean_rating=mean(rating))%>%arrange(desc(mean_rating))%>%slice(1:20)
train%>%group_by(userId)%>%summarize(n=n(),mean_rating=mean(rating))%>%arrange(mean_rating)%>%slice(1:20)
train%>%group_by(userId)%>%summarize(n=n(),mean_rating=mean(rating))%>%ggplot(aes(mean_rating,n)) + geom_point()

```
 
 From the above scatter graph, the distribution of the mean rating of movies per each user is normal.
 
  #Genres
```{r,echo=FALSE}
length(unique(train$genres))
train%>%mutate(genres=factor(genres))%>%group_by(genres)%>%summarize(n=n(),mean_rating=mean(rating))%>%arrange(desc(mean_rating))
train%>%mutate(genres=factor(genres))%>%group_by(genres)%>%summarize(n=n()/1000,mean_rating=mean(rating))%>%ggplot(aes(mean_rating,n)) + geom_point()

```
  There are 797 unique genres.
  The most watched genre is drama.
  There seems to be a genres bias, as the genre with the highest mean_rating(Animation) has only 7 movies.
  Thus, regularization will be needed to remove the bias.
 
   #Timestamp
   
  Split the timestamp to date

```{r,echo=FALSE}
library(lubridate)
train%>%mutate(date=as_datetime(timestamp))%>%mutate(date=round_date(date,unit="week"))%>%group_by(date)%>%summarize(rating=mean(rating))%>%ggplot(aes(date,rating)) + geom_point() + geom_smooth()

```

  From the graph, there is some effect of time on the movie rating, however very small.Thus I will not use "date" as a variable in my model.
  
  # Modelling Approach.
  The variables that have a significant effect on the movie rating are UserId, movieId and genres.
  
# Modelling and Results
  
    # First Model: Just the mean

```{r,echo=FALSE}
mu<-mean(train$rating)
rmse_1<-RMSE(mu,test$rating)
Results<-data.frame(Model="Just the Mean",RMSE=rmse_1)
Results
```

  
  #Second Model:The movie Effect
```{r,echo=FALSE}
movie_effect<-train%>%group_by(movieId)%>%summarize(b_i=mean(rating-mu))
pred_2<-test%>%left_join(movie_effect,by="movieId")%>%mutate(pred=mu+b_i)%>%.$pred
rmse_2<-RMSE(pred_2,test$rating,na.rm = TRUE)
Results<-rbind(Results,data.frame(Model="Movie Effect",RMSE=rmse_2))
Results

```
 #The Third model: The movie + user Effect  
```{r,echo=FALSE}
user_effect<-train%>%left_join(movie_effect,by="movieId")%>%group_by(userId)%>%summarize(b_u=mean(rating-mu-b_i))
pred_3<-test%>%left_join(movie_effect,by="movieId")%>%left_join(user_effect,by="userId")%>%mutate(pred=mu+b_i+b_u)%>%.$pred
rmse_3<-RMSE(pred_3,test$rating,na.rm = TRUE)
Results<-rbind(Results,data.frame(Model="Movie + User Effect",RMSE=rmse_3))
Results
```
 
   #The Forth model: The movie + user  + genre Effect  
```{r,echo=FALSE}
genre_effect<-train%>%left_join(movie_effect,by="movieId")%>%left_join(user_effect,by="userId")%>%group_by(genres)%>%summarize(b_g=mean(rating-mu-b_i-b_u))
test<-test%>%mutate(genres=factor(genres))
pred_4<-test%>%left_join(movie_effect,by="movieId")%>%left_join(user_effect,by="userId")%>%left_join(genre_effect,by="genres")%>%mutate(pred=mu+b_i+b_u+b_g)%>%.$pred
rmse_4<-RMSE(pred_4,test$rating,na.rm = TRUE)
Results<-rbind(Results,data.frame(Model="Movie + User + Genre Effect",RMSE=rmse_4))
Results
```
# The fifth Model: Regularization of the variables

From the above Results, the RMSE keeps decreasing.
From the Analysis we did, we noticed that there is bias in the movieId, userId and genres, due to the few number of ratings in each class. Thus, the next step will be to regularize the variables so as to remove the bias.
The more a movie is rated,the more a user rates movies and the more a certain type of genre is rated,  the better we can recommend a certain type of movie in a specific genre to a specific user.

Regularization involves penalizing the movies that have the least number of ratings, the users who have rated the least number of movies and the genres that have been rated the least number of times.
   
We will need to identify the threshold number that will produce the lowest RMSE.   

```{r,echo=FALSE}
lambda<-seq(0,50,1)
test<-test%>%mutate(genres=factor(genres))
rmse<-function(lambda){
 movie_reg<-train%>%group_by(movieId)%>%summarize(b_i=sum(rating-mu)/(n() +lambda))
 user_reg<-train%>%left_join(movie_reg,by="movieId")%>%group_by(userId)%>%summarize(b_u=sum(rating-mu-b_i)/(n()+lambda))
 genre_reg<-train%>%left_join(movie_reg,by="movieId")%>%left_join(user_reg,by="userId")%>%group_by(genres)%>%summarize(b_g=sum(rating-mu-b_i-b_u)/(n()+lambda))
pred_5<-test%>%left_join(movie_reg,by="movieId")%>%left_join(user_reg,by="userId")%>%left_join(genre_reg,by="genres")%>%mutate(pred=mu+b_i+b_u+b_g)
rmse_5<-RMSE(pred_5$pred,test$rating,na.rm = TRUE)
}
rmses<-data.frame(lambda,sapply(lambda, rmse))
optimal<- rmses[which.min(rmses[,2]),2]#This code may take around 5 minutes
optimal

```
```{r,echo=FALSE}
Results<-rbind(Results,data.frame(Model="Regularization Effect",RMSE=optimal))
Results
```

 Regularization has greatly reduced the RMSE.
 The final model will be that which give the minimum RMSE, which is the Fifth model,i.e, the After the regularization of the variables.
 The value of lambda that produces the minimum RMSE is `r rmses[which.min(rmses[,2]),1]`
 
 
#Test the final model on the Validation data
 
```{r,echo=FALSE}
optimal_lambda <-rmses[which.min(rmses[,2]),1]
movie_reg<-train%>%group_by(movieId)%>%summarize(b_i=sum(rating-mu)/(n() +optimal_lambda))
 user_reg<-train%>%left_join(movie_reg,by="movieId")%>%group_by(userId)%>%summarize(b_u=sum(rating-mu-b_i)/(n()+optimal_lambda))
 genre_reg<-train%>%left_join(movie_reg,by="movieId")%>%left_join(user_reg,by="userId")%>%group_by(genres)%>%summarize(b_g=sum(rating-mu-b_i-b_u)/(n()+optimal_lambda))
final_predictions<-validation%>%left_join(movie_reg,by="movieId")%>%left_join(user_reg,by="userId")%>%left_join(genre_reg,by="genres")%>%mutate(pred=mu+b_i+b_u+b_g)
Final_RMSE<-RMSE(final_predictions$pred,validation$rating,na.rm = TRUE)
Final_RMSE

```

#Conclusion
 An RMSE of `r Final_RMSE` has been attained on the Validation data.
 The model, based on 3 variables can be improved by increasing the number of variables that have an effect on the predicted rating of a movie recommended.
 This model is only limited to the movies that have an existing rating and those that are in the movie site.
 The more a movie is rated the better the predictions, and thus as more users watch and rate the movies, the better the model will become.
 
 Thus, the model has potential to reduce the RMSE and improve predictions in future as more users rate more movies, which will reduce the user and movie bias.
 
 
